### 2010-03-12 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: ispell. Gosh.

\begin{wstructure}
<- The duality of Sigma
    <- Sigma generalizes sum over arbitrary arities
        -> \Sigma A B == \Sigma_{x : A} B x
    <- Sigma generalizes product to have a dependant second component
        -> \Sigma A B == (x : A) \times (B x)
\end{wstructure}

In dependently-typed language, $\Sigma$-types can be read in two
rather different ways. This duality is actually reflected in the
notation we can find in the literature, depending on the sensibility
of the author. Hence, the type $\SigmaTy{A}{B}$ is sometimes written
$\Sigma_{x : A} (B x)$. This notation stresses the fact that
$\Sigma$-types are a generalization of sums over arbitrary
arities. When simply-typed languages have finite sums,
dependently-typed languages have sums of indexed by any set.

On the other hand, $\SigmaTy{A}{B}$ is sometimes written
$\SIGMA{\V{x}}{A} (B x)$. Under this perspective, $\Sigma$-types can
be read as a generalization of products, where the second component
can depend on the first one. When simply-typed languages pack data
into non-dependent tuples, dependently-types languages tolerate this
non-dependent usage. However, they also give the rather novel ability
for data to influence further data.

@@

In the simply-typed world, the essence of data-type definition can be
sumed up by the term ``sum-of-product''. A data-type is defined by a
finite sum, choice, of constructors, each of them composed by a
product, a tuple, of arguments. Therefore, the grammar of data-types
definition in simply-typed languages, such as Haskell or OCaml, follow
this form.

@@

The key idea behind universe construction is our ability to make names
by defining new types. These names are called \emph{codes}. By
defining a set of codes, we somehow define the syntax of a
language. However, as such, a system of code is useless as it lacks a
semantics. Instead of equiping the universe of codes with some
computational behavior, we pragmatically chose to \emph{interpret}
these codes back into the standard type theory. Hence, codes act only
as labels, while the type theory provides the computational machinery.

@@

Finally, we are not limited to one recursive argument. This is
demonstated by our description of a binary tree functor below:

@@

\begin{wstructure}
<- Tagged description
    <- Form TDesc = List (UId x Desc) [equation]
    <- Follow usual sums-of-product presentation of data-type
        <- Finite set of constructors
        <- Then whatever you want
    -> Any Desc data-type can be turned into this form
        -> No loss of expressive power
        /> Garantee a ``constructor form''
\end{wstructure}

@@

So far, for conveniency, we have taken as granted the existence of
$\Desc$, as presented in Figure~\ref{fig:desc_universe}. In
particular, we have considered its code, $\DSigma{}{}$, $\DIndx{}$,
and $\DUnit$ as type formers, extending the basic type theory. Hence,
in an implementation of this type theory, we would have extended it
with these constructors and their typing rules. Although it makes no
conceptual difference, the code of description should rather be read
as a specification. We promise the existence of such objects,
satisfying the typing rules. It will be the subject of
Section~\ref{sec:desc-levitate} to fulfill this promise, by actually
implementing the specification.

@@

\begin{wstructure}
<- Build the fix-point of functors
    <- See examples: need to build their initial algebra
    -> Extend the type theory with Mu/Con [figure]
        <- Straightforward definition of a fixpoint
            <- Interpret D with (Mu D) as subobjects
\end{wstructure}

@@

Let us constructively prove this. First of all, for an object $D$
describing a signature functor, we define its fix-point by the type
former $\Mu{D}$. Inhabitants of the fix-point are materialized by the
structure map: $\Con{x}$ inhabits $\Mu{D}$ if $x$ belongs to the
interpretation of $D$ with the carrier of the initial algebra:

@@

The first step consists in turning $T$ into a predicate
$\Bhab{P}{\Mu{D} \To \Set}$, hence introducing a dependency on the
descriptions. Consequently, we strengthen the algebra $\descop{D}{T}
\To T$ by a similar process. We start with turning the arrow into a
dependenty arrow, hence offering $\Bhab{xs}{\descop{D}{(\Mu{D})}}$ to
further analysis.

Then, we have to translate the notion of algebra in our dependent
setting. This is achieved by the $\All{D}{X}{P}{x}$ predicate that,
intuitively, states that $\Bhab{P}{X \To \Set}$ holds everywhere in
the sub-structures of $x$. Based on that knowledge, we ought to be
able to prove that $P$ holds for $\Con{xs}$. We close our elimination
principle by taking an argument $x$ in $\Mu{D}$ and claiming that the
motive $P$ is realized on $x$.

@@

Whereas simply-typed languages feature a convenient grammar for
defining data-types and manipulating them, our proposal could be seen
as a step backward. In our setting, the user would have to write a
description instead of a sugared data-type definition. Then, to
recognize an element of $\Nat$ for instance, she would have to
identify $\Mu{}$ of a description as indeed the fix-point of the
functor signature of natural numbers. Finally, she would have to write
codes instead of the natural data-type constructor.

As for finite sets previously, these symptoms clearly indicate that we
suffer from too much type information. Just as for finite, the cure
lies in rationalizing this information. Let us first tackle the issue
of data-type declaration: unsurprisingly, the grammar of data-types we
are used can be straighforwardly translated to the
``sigmas-of-sigmas'' paradigm. Hence, a data-type definition is turned
into the corresponding description. Then, at the level of definitions,
we associate the name of the data-type to the fix-point of the
description. The induction principle as well as other generic
operations follow for free.

@@

\note{ This is rather drafty }
\[
\Rule{\forall e_i \in \vec{e}, \propag{\push{e_i}
                                            {\descop{f\: i}{(\Mu{(\DSigma{(\Enum\: b)}{f})})}}}
                                      {e'_i}}
     {\propag{\push{C\: \vec{e}}{\Mu{(\DSigma{(\Enum\: b)}{f})}}}
             {\Con{(\etag{C}, \widetilde{e'})}}}\;\etag{C} \in b
\]

Hence, in the high-level expression language, the user types a
familiar type constructors, applied to a telescope of arguments. In
the type propagation setting, this expression will be pushed to a
$\Mu{}$ type of a tagged description. It is then straightforward to
lookup the tag and turn the constructor form into the appropriate
code.

>>>>

\begin{wstructure}
<- The duality of Sigma
    <- Sigma generalises sum over arbitrary arities
        -> \Sigma A B == \Sigma_{x : A} B x
    <- Sigma generalises product to have a dependant second component
        -> \Sigma A B == (x : A) \times (B x)
\end{wstructure}

In dependently-typed language, $\Sigma$-types can be read in two
rather different ways. This duality is actually reflected in the
notation we can find in the literature, depending on the sensibility
of the author. Hence, the type $\SigmaTy{A}{B}$ is sometimes written
$\Sigma_{x : A} (B x)$. This notation stresses the fact that
$\Sigma$-types are a generalisation of sums over arbitrary
arities. When simply-typed languages have finite sums,
dependently-typed languages have sums of indexed by any set.

On the other hand, $\SigmaTy{A}{B}$ is sometimes written
$\SIGMA{\V{x}}{A} (B x)$. Under this perspective, $\Sigma$-types can
be read as a generalisation of products, where the second component
can depend on the first one. When simply-typed languages pack data
into non-dependent tuples, dependently-types languages tolerate this
non-dependent usage. However, they also give the rather novel ability
for data to influence further data.

@@

In the simply-typed world, the essence of data-type definition can be
summed up by the term ``sum-of-product''. A data-type is defined by a
finite sum, choice, of constructors, each of them composed by a
product, a tuple, of arguments. Therefore, the grammar of data-types
definition in simply-typed languages, such as Haskell or OCaml, follow
this form.

@@

The key idea behind universe construction is our ability to make names
by defining new types. These names are called \emph{codes}. By
defining a set of codes, we somehow define the syntax of a
language. However, as such, a system of code is useless as it lacks a
semantics. Instead of equipping the universe of codes with some
computational behaviour, we pragmatically chose to \emph{interpret}
these codes back into the standard type theory. Hence, codes act only
as labels, while the type theory provides the computational machinery.

@@

Finally, we are not limited to one recursive argument. This is
demonstrated by our description of a binary tree functor below:

@@

\begin{wstructure}
<- Tagged description
    <- Form TDesc = List (UId x Desc) [equation]
    <- Follow usual sums-of-product presentation of data-type
        <- Finite set of constructors
        <- Then whatever you want
    -> Any Desc data-type can be turned into this form
        -> No loss of expressive power
        /> Guarantee a ``constructor form''
\end{wstructure}

@@

So far, for convenience, we have taken as granted the existence of
$\Desc$, as presented in Figure~\ref{fig:desc_universe}. In
particular, we have considered its code, $\DSigma{}{}$, $\DIndx{}$,
and $\DUnit$ as type formers, extending the basic type theory. Hence,
in an implementation of this type theory, we would have extended it
with these constructors and their typing rules. Although it makes no
conceptual difference, the code of description should rather be read
as a specification. We promise the existence of such objects,
satisfying the typing rules. It will be the subject of
Section~\ref{sec:desc-levitate} to fulfil this promise, by actually
implementing the specification.

@@

\begin{wstructure}
<- Build the fix-point of functors
    <- See examples: need to build their initial algebra
    -> Extend the type theory with Mu/Con [figure]
        <- Straightforward definition of a fix-point
            <- Interpret D with (Mu D) as sub-objects
\end{wstructure}

@@

Let us constructively prove this. First of all, for an object $D$
describing a signature functor, we define its fix-point by the type
former $\Mu{D}$. Inhabitants of the fix-point are materialised by the
structure map: $\Con{x}$ inhabits $\Mu{D}$ if $x$ belongs to the
interpretation of $D$ with the carrier of the initial algebra:

@@

The first step consists in turning $T$ into a predicate
$\Bhab{P}{\Mu{D} \To \Set}$, hence introducing a dependency on the
descriptions. Consequently, we strengthen the algebra $\descop{D}{T}
\To T$ by a similar process. We start with turning the arrow into a
dependently arrow, hence offering $\Bhab{xs}{\descop{D}{(\Mu{D})}}$ to
further analysis.

Then, we have to translate the notion of algebra in our dependent
setting. This is achieved by the $\All{D}{X}{P}{x}$ predicate that,
intuitively, states that $\Bhab{P}{X \To \Set}$ holds everywhere in
the sub-structures of $x$. Based on that knowledge, we ought to be
able to prove that $P$ holds for $\Con{xs}$. We close our elimination
principle by taking an argument $x$ in $\Mu{D}$ and claiming that the
motive $P$ is realised on $x$.

@@

Whereas simply-typed languages feature a convenient grammar for
defining data-types and manipulating them, our proposal could be seen
as a step backward. In our setting, the user would have to write a
description instead of a sugared data-type definition. Then, to
recognise an element of $\Nat$ for instance, she would have to
identify $\Mu{}$ of a description as indeed the fix-point of the
functor signature of natural numbers. Finally, she would have to write
codes instead of the natural data-type constructor.

As for finite sets previously, these symptoms clearly indicate that we
suffer from too much type information. Just as for finite, the cure
lies in rationalising this information. Let us first tackle the issue
of data-type declaration: unsurprisingly, the grammar of data-types we
are used can be straightforwardly translated to the
``sigmas-of-sigmas'' paradigm. Hence, a data-type definition is turned
into the corresponding description. Then, at the level of definitions,
we associate the name of the data-type to the fix-point of the
description. The induction principle as well as other generic
operations follow for free.

@@

\note{ This is rather drafty. I should explain what $\vec{e}$ and $\widetilde{e'}$ are. }

\[
\Rule{\forall e_i \in \vec{e}, \propag{\push{e_i}
                                            {\descop{f\: i}{(\Mu{(\DSigma{(\Enum\: b)}{f})})}}}
                                      {e'_i}}
     {\propag{\push{C\: \vec{e}}{\Mu{(\DSigma{(\Enum\: b)}{f})}}}
             {\Con{(\etag{C}, \widetilde{e'})}}}\;\etag{C} \in b
\]

Hence, in the high-level expression language, the user types a
familiar type constructors, applied to a telescope of arguments. In
the type propagation setting, this expression will be pushed to a
$\Mu{}$ type of a tagged description. It is then straightforward to
look-up the tag and turn the constructor form into the appropriate
code.

### 2010-03-12 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: ispell. Gosh.

In this section, we will fulfill several promises. One promise we made
was to effectively implement the code of the $\Desc$ universe. Another
promise we tacitly made is to implement the type former of finite
sets. Despite being a perilous pedagogical exercise for the authors,
we hope to convey to the reader the dizzy feeling of levitation,
without the falling.

@@

The names $\NilE$ and $\ConsE{}{}$ are absolutely not a
coincidence. $\EnumU$ is indeed simply isomorph to a list, with
$\UId$s as parameter. However, in Section~\ref{sec:desc-examples}, we
have shown how to build lists in the universe of
description. Therefore, the actual implementation of $\EnumU$ sums up
to the following, trivial definition:

@@

If not on the content, this modification had an effect on the
names. The type naming scheme of the type theory has condensed: named
type formes ($\EnumU$) and constructors ($\NilE$ and $\ConsE$) are now
replaced by codes and their fix-point in the universe of
descriptions. Our next step is similar, in essence: we are going to
condense the entire naming scheme of data-types \emph{in itself}.

@@

\begin{wstructure}
<- Realizing our promises
    <- We are going to implement Desc
    /> Desc is itself a data-type
        <- Same situation as EnumU
            <- We want to benefit from generic operations
        -> It ought to be encoded in itself
\end{wstructure}

We shall now fulfill our promises. We are going to implement the codes
of the universe of descriptions. Interestingly, in and by itself, a
code is nothing but a data-type. We are in the same situation than
with $\EnumU$: we ought to be able to describe the codes of $\Desc$ in
$\Desc$ itself. Hence, this code would be a first-class citizen, born
with the standard, generic equipment of data-types.

@@

Let us explain how we have proceeded and the obstacle we face. Much as
the data-types we have seen so far, $\DescD$ first consists in chosing
a constructor. Constructors correspond to the code's cases: $\DUnit$,
$\DSigma$, and $\DIndx$. The reader will notice that the tagged
notation we have used so far for $\Desc$ constructors now fully make
sense: these were actually the tags we are defining.

@@

Note that $\DIndx$ is isomorph to $\DHindx{\Unit}$. However, we
tolerate this duplication and do not replace $\DIndx$ by the more
general $\DHindx$. Indeed, unlike its counterpart, $DIndx$ is
first-order. Therefore, while both codes are \emph{extensionally} the
same, in most practical implementation, they would be dealt with
rather differently. While we can definitional equality to cope with
first order objects, the functional presentation introduced by
$\DHindx$ is unlikely to be amenable to a purely definitional
treatment.

Equipped with $\DHindx$, we can describe the extended code of our
universe of data-types, as shown in Figure~\ref{fig:desc-levitate}.
The $\DUnit$ and $\DIndx$ cases remains unchanged, as expected. We
achieve the description of the $\DSigma$ case, by a simple appeal to
the higher-order induction on $S$. The $\DHindx$ case consists in
packing the $H$ in $\Set$ with a recursive arguement.

@@

But this is rather problematic! We are still in the process of
constructing $\DescD$, and the motive of $\F{switch}$ is abruptely
begging for an $\eta$-expanded $\Desc$. Despite our willingness, we
will never be able to materialize such motive.

@@

Before moving on some examples of generic programming, we have to
carry out some maintenance work. The extension of $\Desc$ with
$\DHindx$ have introduced some duplication of functionality. In order
to remove this clutter, we will factor out these commonalities. This
results in a new definition of the $\Desc$ universe. This
reorganization is strictly equivalent to the previous one
(Fig.~\ref{fig:hindx_desc}). Just as the previous version, it is also
self-descriptive. We refer the reader to our companion technical
report for details.

@@

The factorized universe of description is presented in
Figure~\ref{fig:factor-desc}. The modifications are justified as
follow. First, $\DHindx$ have introduced a notion of function space:
we capture this notion by the $\DPi$ constructor. Second, both
$\DHindx$ and $\DIndx$ are composed by a product and the functorial
argument: we capture these two notions with $\DProd{}{}$ and $\DId$,
respectively.

@@

Consequently, the examples developped above can be straightforwardly
translated into this new presentation. \note{We might want to re-do
  the Nat example here, as we encode once again in $\IDesc$} In the
following, we adopt this last version as our de facto universe of
inductive types. In particular, we are going to demonstrate several
instances of generic programming.

@@

\caption{Factorized universe of descriptions}
\label{fig:factor-desc}

@@

The catamorphism can be characterized as an induction on the
description $D$, with a non-dependent motive targetting $T$. Given a
node $xs$ and the induction hypotheses, the method ought to build an
element of $T$. Provided we know how to make an element of
$\descop{D}{T}$, this step will be performed by the algebra $f$. Let
us take a look at this jigsaw:

@@

\note{Update to the factorized presentation}

@@

Of course, we must equip the resulting datatypes with operations
delivering a monadic interface. As usual, \(\LAM{\x}\DVar{\x}\)
performs the r\^ole of ``return'', embedding variables into terms. The
``bind'' operation corresponds to \emph{substitution}. We will now
implement it, as a generic function.

Our implementation will appeal to the $\cata$ function developped
previously. So, let us write down the types, and fill as much
arguments to $\cata$ as possible:

@@

Filling the $???$ subgoal with $\apply\: D\: X\: Y\: \sigma$ finishes
the implementation. To sum up, we have implemented the free monad
construction for an arbitrary tagged description. This gives our user
the ability, for any data-type, to extend it with a notion of
variable. Then, with have equipped this structure with the
corresponding monadic operation, ``bind'' and ``return''. Whereas the
``return'' is a trivial variable former, the ``bind'' operation
corresponds to the substitution of variables by terms. We have shown
how to implement it, for any free monad.

>>>>

In this section, we will fulfil several promises. One promise we made
was to effectively implement the code of the $\Desc$ universe. Another
promise we tacitly made is to implement the type former of finite
sets. Despite being a perilous pedagogical exercise for the authors,
we hope to convey to the reader the dizzy feeling of levitation,
without the falling.

@@

The names $\NilE$ and $\ConsE{}{}$ are absolutely not a
coincidence. $\EnumU$ is indeed simply isomorphic to a list, with
$\UId$s as parameter. However, in Section~\ref{sec:desc-examples}, we
have shown how to build lists in the universe of
description. Therefore, the actual implementation of $\EnumU$ sums up
to the following, trivial definition:

@@

If not on the content, this modification had an effect on the
names. The type naming scheme of the type theory has condensed: named
type formers ($\EnumU$) and constructors ($\NilE$ and $\ConsE$) are now
replaced by codes and their fix-point in the universe of
descriptions. Our next step is similar, in essence: we are going to
condense the entire naming scheme of data-types \emph{in itself}.

@@

\begin{wstructure}
<- Realising our promises
    <- We are going to implement Desc
    /> Desc is itself a data-type
        <- Same situation as EnumU
            <- We want to benefit from generic operations
        -> It ought to be encoded in itself
\end{wstructure}

We shall now fulfil our promises. We are going to implement the codes
of the universe of descriptions. Interestingly, in and by itself, a
code is nothing but a data-type. We are in the same situation than
with $\EnumU$: we ought to be able to describe the codes of $\Desc$ in
$\Desc$ itself. Hence, this code would be a first-class citizen, born
with the standard, generic equipment of data-types.

@@

Let us explain how we have proceeded and the obstacle we face. Much as
the data-types we have seen so far, $\DescD$ first consists in choosing
a constructor. Constructors correspond to the code's cases: $\DUnit$,
$\DSigma$, and $\DIndx$. The reader will notice that the tagged
notation we have used so far for $\Desc$ constructors now fully make
sense: these were actually the tags we are defining.

@@

Note that $\DIndx$ is isomorphic to $\DHindx{\Unit}$. However, we
tolerate this duplication and do not replace $\DIndx$ by the more
general $\DHindx$. Indeed, unlike its counterpart, $DIndx$ is
first-order. Therefore, while both codes are \emph{extensionally} the
same, in most practical implementation, they would be dealt with
rather differently. While we can definitional equality to cope with
first order objects, the functional presentation introduced by
$\DHindx$ is unlikely to be amenable to a purely definitional
treatment.

Equipped with $\DHindx$, we can describe the extended code of our
universe of data-types, as shown in Figure~\ref{fig:desc-levitate}.
The $\DUnit$ and $\DIndx$ cases remains unchanged, as expected. We
achieve the description of the $\DSigma$ case, by a simple appeal to
the higher-order induction on $S$. The $\DHindx$ case consists in
packing the $H$ in $\Set$ with a recursive argument.

@@

But this is rather problematic! We are still in the process of
constructing $\DescD$, and the motive of $\F{switch}$ is abruptly
begging for an $\eta$-expanded $\Desc$. Despite our willingness, we
will never be able to materialise such motive.

@@

Before moving on some examples of generic programming, we have to
carry out some maintenance work. The extension of $\Desc$ with
$\DHindx$ have introduced some duplication of functionality. In order
to remove this clutter, we will factor out these commonalities. This
results in a new definition of the $\Desc$ universe. This
reorganisation is strictly equivalent to the previous one
(Fig.~\ref{fig:hindx_desc}). Just as the previous version, it is also
self-descriptive. We refer the reader to our companion technical
report for details.

@@

The factorised universe of description is presented in
Figure~\ref{fig:factor-desc}. The modifications are justified as
follow. First, $\DHindx$ have introduced a notion of function space:
we capture this notion by the $\DPi$ constructor. Second, both
$\DHindx$ and $\DIndx$ are composed by a product and the functorial
argument: we capture these two notions with $\DProd{}{}$ and $\DId$,
respectively.

@@

Consequently, the examples developed above can be straightforwardly
translated into this new presentation. \note{We might want to re-do
  the Nat example here, as we encode once again in $\IDesc$} In the
following, we adopt this last version as our de facto universe of
inductive types. In particular, we are going to demonstrate several
instances of generic programming.

@@

\caption{Factorised universe of descriptions}
\label{fig:factor-desc}

@@

The catamorphism can be characterised as an induction on the
description $D$, with a non-dependent motive targeting $T$. Given a
node $xs$ and the induction hypotheses, the method ought to build an
element of $T$. Provided we know how to make an element of
$\descop{D}{T}$, this step will be performed by the algebra $f$. Let
us take a look at this jigsaw:

@@

\note{Update to the factorised presentation}

@@

Of course, we must equip the resulting data-types with operations
delivering a monadic interface. As usual, \(\LAM{\x}\DVar{\x}\)
performs the r\^ole of ``return'', embedding variables into terms. The
``bind'' operation corresponds to \emph{substitution}. We will now
implement it, as a generic function.

Our implementation will appeal to the $\cata$ function developed
previously. So, let us write down the types, and fill as much
arguments to $\cata$ as possible:

@@

Filling the $???$ sub-goal with $\apply\: D\: X\: Y\: \sigma$ finishes
the implementation. To sum up, we have implemented the free monad
construction for an arbitrary tagged description. This gives our user
the ability, for any data-type, to extend it with a notion of
variable. Then, with have equipped this structure with the
corresponding monadic operation, ``bind'' and ``return''. Whereas the
``return'' is a trivial variable former, the ``bind'' operation
corresponds to the substitution of variables by terms. We have shown
how to implement it, for any free monad.

### 2010-03-12 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: more philosophical questions
Desc@ICFP: fix functor for lambda terms
Desc@ICFP: ispell. Gosh.

This generalization can be achieved by a minor evolution of $\Desc$
(Fig.~\ref{fig:factor-desc}). Indeed, the functorial nature of $\Desc$
is entirely captured by the $\DId$ code. We get an indexed functor
by turning $\DId$ into an index-dependent $\DVar$ code:

@@

@@

Because $\Nat$ is just an inductive type, we make no use of the index:
it is the trivial inhabitant of $\Unit$. Therefore, the recursive
argument is materialized by $\DVar{\Void}$, where we were using $\DId$
in the $\Desc$ presentation. This transformation applies to all
inductive types: we have not lost in expressive power during the
transition.

@@

The $\Fin$ example is also a well-known example of Generalized
Abstract Data-Type (GADT), in Haskell. It comes as no surprise that
our universe of indexed description supports this restricted class of
dependent data-types. As we have shown with $\Fin$, constrained
constructors are straightforwardly translated into equality
constraints.

@@

\begin{wstructure}
!!! Need Help !!!
<- Brady optimization: forcing
    <- Source to source translation
    <- Able to remove some constraints
    -> Example: Fin [figure]
    ??? More technical detail needed
\end{wstructure}

To get a clue of a solution, let us look back at $\Fin$. We note that
the equation is introduced because we are \emph{storing} the index of
the inductive family, through $m$. Hence the constraint. However,
\emph{inductive families need not store their
  indices}~\cite{brady:index-inductive-families}. Applying the
\emph{forcing} optimization to our definition of $\Fin$ gives the
following, equivalent definition:

@@

Equations have simply disappeared! We should precise that forcing a
description is not garanteed to remove all constraints. It is subject
of future work to see if constraints can be entirely eradicated, or
presented more conveniently to the user.

Finally, it is worth mentioning that the forcing optimization is a
\emph{source-to-source} transformation of the description.

@@

\begin{wstructure}
<- Variation on a theme: free imonad construction
    <- Recall existence of generic free monad construction
    -> Present its generalization to IDesc [equation]
        <- \I -> IDesc I as describing an indexed endofunctor
        <- Free monad construction
    -> Still a suitable, generic notion of substitution
        <- show type signature
        <- show implementation?? (space! space!)
\end{wstructure}

@@

\[\stk{
\LambdaTFreeD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}ll}
\LambdaTFreeD\: n = \DSigma{}{} & (\EnumT [ \LambdaTApp, \LambdaTLam ]) \\
                                & \left[\begin{array}{l}
                                  \DProd{\DVar{n}}{\DVar{n}} \\
                                  \DVar{(\NatSuc{n})}
                                  \end{array}\right]
\end{array}
}\]

>>>>

This generalisation can be achieved by a minor evolution of $\Desc$
(Fig.~\ref{fig:factor-desc}). Indeed, the functorial nature of $\Desc$
is entirely captured by the $\DId$ code. We get an indexed functor
by turning $\DId$ into an index-dependent $\DVar$ code:

@@

\note{Should we say something about IMu, fixed-points, and their
  elimination before moving on? If so, where should we put it?}

@@

Because $\Nat$ is just an inductive type, we make no use of the index:
it is the trivial inhabitant of $\Unit$. Therefore, the recursive
argument is materialised by $\DVar{\Void}$, where we were using $\DId$
in the $\Desc$ presentation. This transformation applies to all
inductive types: we have not lost in expressive power during the
transition.

@@

The $\Fin$ example is also a well-known example of Generalised
Abstract Data-Type (GADT), in Haskell. It comes as no surprise that
our universe of indexed description supports this restricted class of
dependent data-types. As we have shown with $\Fin$, constrained
constructors are straightforwardly translated into equality
constraints.

@@

\begin{wstructure}
!!! Need Help !!!
<- Brady optimisation: forcing
    <- Source to source translation
    <- Able to remove some constraints
    -> Example: Fin [figure]
    ??? More technical detail needed
\end{wstructure}

To get a clue of a solution, let us look back at $\Fin$. We note that
the equation is introduced because we are \emph{storing} the index of
the inductive family, through $m$. Hence the constraint. However,
\emph{inductive families need not store their
  indices}~\cite{brady:index-inductive-families}. Applying the
\emph{forcing} optimisation to our definition of $\Fin$ gives the
following, equivalent definition:

@@

Equations have simply disappeared! We should precise that forcing a
description is not guaranteed to remove all constraints. It is subject
of future work to see if constraints can be entirely eradicated, or
presented more conveniently to the user.

Finally, it is worth mentioning that the forcing optimisation is a
\emph{source-to-source} transformation of the description.

@@

\begin{wstructure}
<- Variation on a theme: free imonad construction
    <- Recall existence of generic free monad construction
    -> Present its generalisation to IDesc [equation]
        <- \I -> IDesc I as describing an indexed endofunctor
        <- Free monad construction
    -> Still a suitable, generic notion of substitution
        <- show type signature
        <- show implementation?? (space! space!)
\end{wstructure}

@@

\[\stk{
\LambdaTFreeD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}ll}
\LambdaTFreeD\: n = \DSigma{}{} & (\EnumT [ \LambdaTApp, \LambdaTLam ]) \\
                                & \left[\begin{array}{l}
                                  \DProd{\DVar{\NatSuc{n}}}{\DVar{n}} \\
                                  \DVar{(\NatSuc{n})}
                                  \end{array}\right]
\end{array}
}\]
